Pyg4ometry: a python library for the creation of Monte Carlo particle transport physical geometries

### 2019-08-07 public/paper/pyg4ometry.tex
Started citations and updated verb/it of pyg4ometry

\title{pyg4ometry: a python library for the creation of Monte Carlo particle transport geometries}

@@

%% main text
\section{Introduction} \label{sec:introduction}
There are numerous different software codes to simulate the passage of particles through material, so called particle transport (PT) programmes, these 
include MCMP, FLUKA, Geant3 and Geant4. All these codes are based on the Monte Carlo technique but each code either has a particular specialism 
or target user community.  PT Monte Carlo simulations have diverse uses, shielding calculations for radiological protection, detector performance, medical 
imaging and therapy, space radiation environment simulation are some examples. A fundamental requirement of all of the codes is to supply a computer 
readable description of the physical geometry that the particles are passing through.  The creation of geometry files is typically a very time consuming activity, 
the simulation validity and performance is directly dependent on the quality of the geometry. The requirement of a  geometry system that allows the conversion 
between files prepared for different codes will enable cross checks of the physics processes in particle transport codes. Parallel to the creation of geometry files 
for PT programs, usually computer assisted design (CAD) files exist for systems which need to be simulated in PT programmes. The fundamental geometric 
representations in CAD files is typically not amenable for PT programmes.  For these reasons it is advantageous to create a software tool that allows particle 
transport code users to rapidly develop geometry files, convert between common PT programme geometry formats and finally import CAD models. This paper 
describes such geometry creation and conversion package called {\em pyg4ometry}, implemented in {\em Python} and internally based on the Geant4 geometry 
description. The main features of  {\em pyg4ometry} are a python scripting API to rapidly design parametrised geometry, conversion to and from  FLUKA geometry 
descriptions, conversion from CAD formats  (STEP and IGES) based on FreeCAD-OpenCascade and powerful geometry visualisation tools based on VTK. 
The origin of pyg4ometry was a set of utilities to prepare geometry for an accelerator beam line simulation based on Geant4 called BDSIM. Accelerator physicists 
like specialists in other areas need a tool to quickly simulate beam losses, and the subsequent interaction of the charged particle beam with the accelerator material.

@@

\subsection{Expression engine}
In GDML symbolic expressions can be used to parametrise solids and their placement. These expressions are evaluated when the GDML is loaded into Geant4. 
In order to fully replicate the functionality of GDML an expression engine was implemented using ANTLR. GDML allows for the definition and assignment of 
variables. GDML expressions are not much more complicated than binary operators $+, -, \times, /$ and common trigonometric and special functions $\sin, \cos, \tan$ etc. 
The AST terminates on either expressions which evaluate to numbers or GDML variables. Internally all classes use GDML expressions and not floating point numbers. 
Storing internal data as expressions allows for deferred evaluation (or re-evaluation) of solid parameters and placements. This allows a user to update variables whilst
defining geometry and the expression engine will update all internal values.

\subsection{Tessellation engine}
For each Geant4 primitive solid a triangular tessellated mesh is generated and cached. This mesh is then used to determine the extent of placed instances of 
geometry (physical volumes) and meshes for CSG derived solids. CSG mesh calculations are performed using a BSP tree technique. Triangular meshes based
on CSG operations involving curved surfaces often contain large numbers of thin triangles. Before meshes are visualised or written to disk various algorithms from 
TetGEN can be employed to give the meshes more desirable features.

@@

\subsection{Input and output} 
For each input format available in {\it pyg4ometry} a dedicated {\it Reader} class is implemented, so \verb|gdml.Reader|, \verb|stl.Reader|, \verb|fluka.Reader| 
and \verb|freecad.Reader|. Each reader constructs the appropriate Geant4 classes and provides a registry which can be used or manipulated by the user. 
Output consists of taking the registry and writing the appropriate GDML.

\subsection{Overlap detection}
All PT transport codes can not handle spatial overlap between two geometric objects and likely are to have ill defined behaviour when tracking particles  
throughsuch a situation.  A key feature of {\it pyg4ometry} is the detection of potential overlaps efficiently, it does this by performing a CSG intersection between solid instances 
and determining if the resulting mesh is empty. Generally the overlap detection algorithm proceeds as

@@

\section{Performance of converted geometry in Geant4}
Clearly for Geant4 geometry defined in {\it pyg4ometry} the performance is exactly the same as it the geometry was specified in C++ and linked against the Geant4
libraries. A comparison of the performance of converted CAD or FLUKA geometry compared with native Geant4 solids is required so a user can optimise the available
computing resources against the human resources (required to convert geometry). BDSIM is used to load the pyg4ometry generated files  and run basic simulations where 
high energy particles are

\section{Conclusions}
The authors believe that tools to quickly create, either ab-initio or by conversion geometry usable in particle transport Monte Carlo programmes will save significant 
amounts of user effort and ultimately yield more accurate simulations. {\it pyg4ometry} is a relatively complete implementation of a geometry creation tool, whilst heavily 
internally based on Geant4 and GDML can have utility for users of all PT codes. {\it pyg4ometry} can clearly be extended to other formats or applications. As for applications,
examples include  the  generation triangular mesh structures for GPU accelerated photon tracking in liquid noble  dark matter detectors and geometries for event visualisation. 
Although not implemented in the current version, a C++ output writer can be quickly implemented to generate geometry which can be compiled into a Geant4 application.

@@

\bibliographystyle{elsarticle-num}
\bibliography{<your-bib-database>}

%% Authors are advised to submit their bibtex database files. They are
%% requested to list a bibtex style file in the manuscript if they do
%% not want to use elsarticle-num.bst.

>>>>

\title{Pyg4ometry: a python library for the creation of Monte Carlo particle transport physical geometries}

@@

%% main text
\section{Introduction} \label{sec:introduction}
There are numerous different software codes to simulate the passage of particles through material, so called particle transport (PT) programmes, these 
include MCNP \cite{Mcnp_Werner}, FLUKA \cite{Fluka_Ferrari,Fluka_Bohlen}, Geant3 \cite{Geant3_Brun} and Geant4 \cite{Geant4_Agostinelli}. 
All these codes are based on the Monte Carlo technique but each code either has a particular specialism or target user community.  PT Monte Carlo simulations 
have diverse uses, shielding calculations for radiological protection, detector performance, medical imaging and therapy and space radiation environment 
simulation are some examples. A fundamental requirement of all of the codes is to supply a computer readable description of the physical geometry that 
the particles are passing through.  The creation of geometry files is typically a very time consuming activity and the simulation validity and performance is directly 
dependent on the quality of the geometry. The requirement of a geometry system that allows the conversion between files prepared for different codes will 
enable cross checks of the physics processes in different particle transport codes. Parallel to the creation of geometry files for PT programs, usually computer assisted 
design (CAD) files exist for systems which need to be simulated in PT programmes. The fundamental geometric representations in CAD files is typically not 
amenable for PT programmes.  For these reasons it is advantageous to create a software tool that allows particle transport code users to rapidly develop 
geometry files, convert between common PT programme geometry formats and finally import CAD models.

This paper describes such geometry creation and conversion package called {\em pyg4ometry}, implemented in {\em Python} and internally based on the Geant4 geometry 
description. The main features of  {\em pyg4ometry} are a python scripting API to rapidly design parametrised geometry, conversion to and from  FLUKA geometry 
descriptions, conversion from CAD formats  (STEP and IGES) based on FreeCAD \cite{FreeCAD} OpenCascade \cite{OpenCASCADE} and powerful geometry 
visualisation tools based on VTK \cite{VTK4}. The origin of pyg4ometry was a set of utilities to prepare geometry for an accelerator beam line simulation based on Geant4 
called BDSIM. Accelerator physicists like specialists in other areas need a tool to quickly simulate beam losses, and the subsequent interaction of the charged particle 
beam with the accelerator material.

There are existing codes that have functionality similar to pyg4ometry. Tools exist to convert CAD files to GDML, e.g. GUIMesh \cite{GUIMesh_Pinto}.
CAD software can export data to STL. The existent set of software does not provide a complete set of tools to efficiently create complex geometries.

@@

\subsection{Expression evaluation}
In GDML symbolic expressions can be used to parametrise solids and their placement. These expressions are evaluated when the GDML is loaded into Geant4. 
In order to fully replicate the functionality of GDML an expression engine was implemented using ANTLR. GDML allows for the definition and assignment of 
variables. GDML expressions are not much more complicated than binary operators $+, -, \times, /$ and common trigonometric and special functions $\sin, \cos, \tan$ etc. 
The AST terminates on either expressions which evaluate to numbers or GDML variables. Internally all classes use GDML expressions and not floating point numbers. 
Storing internal data as expressions allows for deferred evaluation (or re-evaluation) of solid parameters and placements. This allows a user to update variables whilst
defining geometry and the expression engine will update all internal values.

\subsection{Tessellation of solids}
For each Geant4 primitive solid a triangular tessellated mesh is generated and cached. This mesh is then used to determine the extent of placed instances of 
geometry (physical volumes) and meshes for CSG derived solids. CSG mesh calculations are performed using a BSP tree technique. Triangular meshes based
on CSG operations involving curved surfaces often contain large numbers of thin triangles. Before meshes are visualised or written to disk various algorithms from 
TetGEN can be employed to give the meshes more desirable features.

@@

\subsection{Input and output} 
For each input format available in pyg4ometry a dedicated {\it Reader} class is implemented, so \verb|gdml.Reader|, \verb|stl.Reader|, \verb|fluka.Reader| 
and \verb|freecad.Reader|. Each reader constructs the appropriate Geant4 classes and provides a registry which can be used or manipulated by the user. 
Output consists of taking the registry and writing the appropriate GDML.

\subsection{Overlap detection}
All PT transport codes can not handle spatial overlap between two geometric objects and likely are to have ill defined behaviour when tracking particles  
throughsuch a situation.  A key feature of pyg4ometry is the detection of potential overlaps efficiently, it does this by performing a CSG intersection between solid instances 
and determining if the resulting mesh is empty. Generally the overlap detection algorithm proceeds as

@@

\section{Performance of converted geometry in Geant4}
Clearly for Geant4 geometry defined in pyg4ometry the performance is exactly the same as it the geometry was specified in C++ and linked against the Geant4
libraries. A comparison of the performance of converted CAD or FLUKA geometry compared with native Geant4 solids is required so a user can optimise the available
computing resources against the human resources (required to convert geometry). BDSIM is used to load the pyg4ometry generated files  and run basic simulations where 
high energy particles are

\section{Conclusions}
The authors believe that tools to quickly create, either ab-initio or by conversion geometry usable in particle transport Monte Carlo programmes will save significant 
amounts of user effort and ultimately yield more accurate simulations. Pyg4ometry is a relatively complete implementation of a geometry creation tool, whilst heavily 
internally based on Geant4 and GDML can have utility for users of all PT codes. Pyg4ometry can clearly be extended to other formats or applications. As for applications,
examples include  the  generation triangular mesh structures for GPU accelerated photon tracking in liquid noble  dark matter detectors and geometries for event visualisation. 
Although not implemented in the current version, a C++ output writer can be quickly implemented to generate geometry which can be compiled into a Geant4 application.

@@

\bibliographystyle{elsarticle-num}
\bibliography{pyg4ometry}
%% Authors are advised to submit their bibtex database files. They are
%% requested to list a bibtex style file in the manuscript if they do
%% not want to use elsarticle-num.bst.
